### 2014-03-19 reanalysis/Fadi_Towfic_generic_medicine/towfic_et_al.rmd
download data

```{r, results='hide', echo=FALSE}
starttime = Sys.time()
```
`r as.character(starttime)`

@@

@@

Table S1 does have a "Chip"- coloumn, unfortunatly there is no dedicated "treatment"-coloumn.
Communication with the corresponding author yielded this explanation
> ..and the only thing we used as a covariate was the unique treatment names themselves (e.g. ³Medium² or ³RS").

@@

```{r}
sampleannotation = read.table("data/sampleannotation.csv", sep="\t", header=TRUE,  stringsAsFactors=FALSE)
sampleannotation$code = make.names(sampleannotation$code)
sampleannotation$chip = as.character(sampleannotation$chip)
dimnames(sampleannotation)[[1]] = sampleannotation$code
head(sampleannotation)
```
The "covariate"-coloumn is made based on the "code"-coloumn, but might not match  to what were actually used for all samples. The last two coloumns are from the GEO deposit and reveal that 3 of the samples do not have data and are taken out.

The nameing convention seeme to differ slightly between the Table S1 and the text in Towfic et al. This is our interpretation of covariate labels and its corresponding name in the text.
- **DP** refferd to as GA (but not GA as in table S1)
- **N** refferd to as "generic"
- **M** reffered to as "medium"
- **RS** reffered to as "reference standard""

@@

The processed data from the above GEO deposit was from another publication which did not processes the data as described in Towfic et al. (personal communication). The lesser processed linked to as [GSE40566_non_normalized.txt.gz](http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file&file=GSE40566%5Fnon%5Fnormalized%2Etxt%2Egz) in the supplementary data was thus used in this report, although it is not confirmed that this is the version of the data actually used in Towfic et al.

Probe annotation was also found in the supplementary data as [GSE40566_RAW.tar](http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file).

A recap of the important input files used in this report:
- **sampleannotation.csv** holds the important sample/batch/covariate assignment. Compiled based on inforamtion in Towfic et al, the geo deposit and personal communication.
- **GSE40566_non_normalized.txt** the measurments in a sample vs probe matrix. From the [GEO deposit](http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file&file=GSE40566%5Fnon%5Fnormalized%2Etxt%2Egz).
- **GPL6887_MouseWG-6_V2_0_R3_11278593_A.txt** the probe annotation needed for the datarow to probe to gene matching. From the GEO deposit inside the [GSE40566_RAW.tar](http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file).

Reading the data.
```{r}
# the output from Illumina software.
rawdata = read.table("not_in_github/GSE40566_non_normalized.txt", sep="\t", header=TRUE, stringsAsFactors=FALSE)

# the probe annotation file found inside GSE40566_RAW.tar
rawannotation = read.table("not_in_github/GPL6887_MouseWG-6_V2_0_R3_11278593_A.txt", sep="\t", header=TRUE, stringsAsFactors=FALSE, skip=8, comment.char="", quote="", fill=TRUE)
```

Some formatting is needed to match the probe information to the measurments and the measurments to the sample information.
```{r}
# annotation file has two separate tables, one for experimental and one at the end for controls.
# splitting and fixing the tables:
tmp =  rawannotation$Species=="Mus musculus"
experimentalannot = rawannotation[tmp,]
experimentalannot$Array_Address_Id = as.numeric(experimentalannot$Array_Address_Id)
controlannot = rawannotation[!tmp,]
dimnames(controlannot)[[2]] = rawannotation[rawannotation[,2]=="Array_Address_Id",]
controlannot$Array_Address_Id = suppressWarnings(as.numeric(controlannot$Array_Address_Id))
controlannot = controlannot[!is.na(controlannot$Array_Address_Id),]
controlannot=controlannot[,1:6]
probeannotation = merge(experimentalannot, controlannot, all=TRUE )
#dim(probeannotation)
rm(tmp, experimentalannot, controlannot)

@@

Several tables with results are presented in the Supporting Information in Towfic et al. We are aiming to reproduce two of these. Unfortunatly they were only presented in a pdf-format not easely parsed by a computer. We had to resort to a ad-hoc method of cut-and-paste from pdf into text files which were somewhat polluted by pdf-formatting code. For some tables about 10% of the rows will be lost, but the rest will suffice for our purpose.

@@

Towfic et al. performed many different test on the data and it proved difficult and to labouros to reproduce all of their results without any source code. Insted we foucus on the key part of testing for differentially expressed genes between "GA" (DP) and "generic"(N) as described in Table S2 and Table S5. But before those tests we have to pre-process according to the description.

@@

```{r reproduced_tables2, dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}
# scatter plot correlation of the 4 columns in table_S2 vs. the reproduced numbers. String correlation.
par(mfrow=c(2, 2))
plot(table_s2$AVG_Medium, meanM[table_s2$ID], pch=".")
plot(table_s2$AVG_Reference_Standard, meanRS[table_s2$ID], pch=".")
plot(table_s2$AVG_GA, meanDP[table_s2$ID], pch=".")
plot(table_s2$AVG_generic, meanN[table_s2$ID], pch=".")
```
The numbers in Table S2 is almost reproduced.

@@

There must be a typo in the figure text, they ment "Comparison of expression in GA to expression in **generic** for each probe, ...".

@@

# plot
plot(table_s5$Fold_Change, foldchange_DP_N[table_s5$Probe],
     ylim=lim, xlim=lim, pch=".", xlab="Table_S5", ylab="reproduced",
     main="FC comparison")
rm(lim)
```
The plot reveals that the reproduced fold changes are much more compressed than the ones in Table S5.
There seems to two linear relationship, one for neagtive FC's and another for positve FC's. They seems to have different intercepts but comparable slope.

@@

The T-test is stright forward.
```{r dev='png', fig.width=10, fig.height=6, fig.show='hold', tidy=FALSE}
ttest_p = apply(datamatrices[["real_combat_covariates"]], 1, 
                function(x) { t.test(x[sampleannotation$covariate=="DP"], 
                                     x[sampleannotation$covariate=="N"])$p.value } )
par(mfrow=c(1, 2))
plot(table_s5[,"t_test_p"], ttest_p[table_s5$Probe], xlab="Original Table S5", 
     ylab="Reproduced", main="t-test P-values from table S5 vs. reproduced")
thiscolors = c("black", "red")
hist(ttest_p, breaks=100, freq=T, border=thiscolors[1], main="Reproduced t-test p-values for DP vs. N")
hist(ttest_p[table_s5$Probe], breaks=100, add=T, freq=T, border=thiscolors[2])
legend("topright", legend=c("All probes", "Table S5 probes"), text.col=thiscolors)
```

@@

The T-test can not handle batch inside the test. An alternative way of performing a t-test is to subset the data and do the comparison whithin each batch. For each of the 7 batches that have both a DP ("GA") and a N ("generic") sample, we can do a direct comparison and in effect ending up with 7 samples. Only 7 samples will give less power but the caveats because of batch effect and adjustments are also far less.
```{r tidy=FALSE, results='hold'}

>>>>

`r as.character(Sys.time())`

@@

```{r runparameters, results='hide', echo=FALSE}
starttime = Sys.time()
debug = FALSE
downloaddata=TRUE
```

@@

Table S1 does have a "Chip"- column, unfortunately there is no dedicated "treatment"-column.
Communication with the corresponding author yielded this explanation
> ..and the only thing we used as a covariate was the unique treatment names themselves (e.g. "Medium"" or "RS").

@@

```{r}
sampleannotation = read.table("data/sampleannotation.csv", sep="\t", header=TRUE,  stringsAsFactors=FALSE)
sampleannotation$code = make.names(sampleannotation$code)
sampleannotation$chip = as.character(sampleannotation$chip)
dimnames(sampleannotation)[[1]] = sampleannotation$code
head(sampleannotation)
```
The "covariate"-column is made based on the "code"-column, but might not match  to what were actually used for all samples. The last two columns are from the GEO deposit and reveal that 3 of the samples do not have data and will be removed from the sampleannotation.

The naming convention seem to differ slightly between the Table S1 and the text in Towfic et al. This is our interpretation of covariate labels and its corresponding name in the text.
- **DP** referred to as GA (but not GA as in table S1)
- **N** referred to as "generic"
- **M** referred to as "medium"
- **RS** referred to as "reference standard""

@@

The processed data from the above GEO deposit ("Series Matrix File") is from another publication which did not processes the data as described in Towfic et al. (personal communication). The lesser processed data in GEO linked to as [GSE40566_non_normalized.txt.gz](http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file&file=GSE40566%5Fnon%5Fnormalized%2Etxt%2Egz) in the "Supplementary file" section was thus used in this report.

Probe annotation was also found in the "Supplementary file" section as [GSE40566_RAW.tar](http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file).

A recap of the important input files used in this report:
- **sampleannotation.csv** holds the important sample/batch/covariate assignment. Compiled based on information in Towfic et al, the geo deposit and personal communication.
- **GSE40566_non_normalized.txt** the measurements in a sample vs probe matrix. From the [GEO deposit](http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file&file=GSE40566%5Fnon%5Fnormalized%2Etxt%2Egz). Not in github.
- **GPL6887_MouseWG-6_V2_0_R3_11278593_A.txt** the probe annotation needed for the data row to probe to gene matching. From the GEO deposit inside the [GSE40566_RAW.tar](http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file). Not in github.

Reading the data.
```{r}

if(downloaddata)
{
  temp <- tempfile()
  download.file(url="http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file&file=GSE40566%5Fnon%5Fnormalized%2Etxt%2Egz",
              destfile=temp, mode = "wb")
  rawdata <- read.table(temp, sep="\t", header=TRUE, stringsAsFactors=FALSE)
  unlink(temp)

  temp <- tempfile()
  download.file(url="http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE40566&format=file",
              destfile=temp, mode = "wb")
  tardirtemp = tempdir()
  untar(temp, exdir = tardirtemp)
  rawannotation = read.table(paste(tardirtemp ,"/GPL6887_MouseWG-6_V2_0_R3_11278593_A.txt.gz", sep=""), 
                             sep="\t", header=TRUE, stringsAsFactors=FALSE, 
                             skip=8, comment.char="", quote="", fill=TRUE)

  unlink(temp)
  unlink(tardirtemp, recursive=TRUE )
}else{

  # if download did not work change downloaddata to FALSE
  # download data from the GEO deposit
  # unpack and place files in a folder named "not_in_github"
  rawdata = read.table("not_in_github/GSE40566_non_normalized.txt", 
                       sep="\t", header=TRUE, stringsAsFactors=FALSE)

  # the probe annotation file found inside GSE40566_RAW.tar
  rawannotation = read.table("not_in_github/GPL6887_MouseWG-6_V2_0_R3_11278593_A.txt", 
                             sep="\t", header=TRUE, stringsAsFactors=FALSE, 
                             skip=8, comment.char="", quote="", fill=TRUE)
}
```

Some formatting is needed to match the probe information to the measurements and the measurements to the sample information.
```{r}
# annotation file has two separate tables, one for experimental and one at the end for controls.
# splitting and fixing the tables:
tmp =  rawannotation$Species=="Mus musculus"
experimentalannot = rawannotation[tmp,]
experimentalannot$Array_Address_Id = as.numeric(experimentalannot$Array_Address_Id)
controlannot = rawannotation[!tmp,]
dimnames(controlannot)[[2]] = rawannotation[rawannotation[,2]=="Array_Address_Id",]
controlannot$Array_Address_Id = suppressWarnings(as.numeric(controlannot$Array_Address_Id))
controlannot = controlannot[!is.na(controlannot$Array_Address_Id),]
controlannot=controlannot[,1:6]
probeannotation = merge(experimentalannot, controlannot, all=TRUE )
#dim(probeannotation)
rm(tmp, experimentalannot, controlannot)

@@

Several tables with results are presented in the Supporting Information in Towfic et al. We are aiming to reproduce two of these. Unfortunately they were only presented in a pdf-format not easily parsed by a computer. We had to resort to a adhoc method of cut-and-paste from pdf into text files which were somewhat polluted by pdf-formatting code. For some tables about 10% of the rows will be lost, but the rest will suffice for our purpose.

@@

Towfic et al. performed many different test on the data and it is outside the scope of this report to reproduce all of their results. We focus on the key part of testing for differentially expressed genes between "GA" (DP) and "generic"(N) as described in Table S2 and Table S5. But before those tests we have to preprocess according to the description.

@@

```{r reproduced_tables2, dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}
# scatter plot correlation of the 4 columns in table_S2 vs. the reproduced numbers. String correlation.
par(mfrow=c(2, 2))
plot(table_s2$AVG_Medium, meanM[table_s2$ID], pch=".")
plot(table_s2$AVG_Reference_Standard, meanRS[table_s2$ID], pch=".")
plot(table_s2$AVG_GA, meanDP[table_s2$ID], pch=".")
plot(table_s2$AVG_generic, meanN[table_s2$ID], pch=".")
```
Original values from table S2 on X-axis, reporduced values on Y-axis. The values in Table S2 is almost reproduced.

@@

There must be a typo in the figure text, they meant "Comparison of expression in GA to expression in **generic** for each probe, ...".

@@

# plot
plot(table_s5$Fold_Change, foldchange_DP_N[table_s5$Probe],
     ylim=lim, xlim=lim, pch=".", xlab="Table_S5", ylab="reproduced",
     main="FC comparison")
rm(lim)
```
The plot reveals that the reproduced fold changes are much more compressed than the ones in Table S5.
There seems to two linear relationship, one for negative FC's and another for positive FC's. They seems to have different intercepts but comparable slope.

@@

The T-test is straight forward.
```{r dev='png', fig.width=10, fig.height=6, fig.show='hold', tidy=FALSE}
ttest_p = apply(datamatrices[["real_combat_covariates"]], 1, 
                function(x) { t.test(x[sampleannotation$covariate=="DP"], 
                                     x[sampleannotation$covariate=="N"])$p.value } )
par(mfrow=c(1, 2))
plot(table_s5[,"t_test_p"], ttest_p[table_s5$Probe], xlab="Original Table S5", 
     ylab="Reproduced", main="t-test P-values from table S5 vs. reproduced")
thiscolors = c("black", "red")
hist(ttest_p, breaks=100, freq=T, border=thiscolors[1], main="Reproduced t-test p-values for DP vs. N")
hist(ttest_p[table_s5$Probe], breaks=100, add=T, freq=T, border=thiscolors[2])
legend("topright", legend=c("All probes", "Table S5 probes"), text.col=thiscolors)
```

@@

The T-test can not handle batch inside the test. An alternative way of performing a t-test is to subset the data and do the comparison within each batch. For each of the 7 batches that have both a DP ("GA") and a N ("generic") sample, we can do a direct comparison and in effect ending up with 7 samples. Only 7 samples will give less power but the caveats because of batch effect and adjustments are also far less.
```{r tidy=FALSE, results='hold'}
